---
title: "Creating a pool set for matchRanges"
author: "Eric S. Davis"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
bibliography: library.bib
output:
  rmarkdown::html_document:
    highlight: tango
    toc: true
    toc_float: true
    fig_width: 5
    fig_height: 3
vignette: |
  %\VignetteIndexEntry{1. Creating a pool set for matchRanges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

`matchRanges()` performs subset selection on a pool of ranges
such that chosen covariates are distributionally matched to 
a focal set of ranges. However, the generation of a set of 
pool ranges is not a trivial task. This vignette provides some
guidance for how to generate a pool of ranges.

For this analysis, we use DNase peaks as a measure of open
chromatin, and ChIP-seq for JUN peaks as a measure of JUN
binding. Suppose we are interested in the properties of
chromatin accessibility, but suspect that JUN-binding impacts
accessibility. We can use `matchRanges()` to control for
the DNase signal and the length of the site so we can compare
our JUN-bound sites (i.e., our `focal` set) to sites where
JUN is not bound (i.e., our `pool` set).

## Obtaining example data

First, we use `AnnotationHub` to access `GRanges` for DNase and
JUN narrowPeaks in human (hg19) GM12878 cells:

```{r, message=FALSE, warning=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()

dnase <- ah[["AH30743"]]
junPeaks <- ah[["AH22814"]]

dnase
```

Since we want to control for accessibility, we can use the
`signalValue` from the DNase peaks as a covariate. DNase sites
are also different lengths. If we suspect length might impact
accessibility differently at JUN-bound sites, we can include it
as a covariate as well. For visualization, let's convert these 
to log-scale using `mutate()` from `plyranges`:

```{r, message=FALSE, warning=FALSE}
library(plyranges)
dnase <- dnase |>
  mutate(log10_signal = log10(signalValue + 1),
         log10_length = log10(width(dnase) + 1))
```

## Creating the `focal` and `pool` sets

Next we define our focal and pool sets. The `focal` set contains
the feature of interest (i.e., DNase peaks bound by JUN),
whereas the `pool` set lacks this feature (i.e., unbound DNase
peaks). `matchRanges()` is designed to handle datasets that can
be binarized into these two distinct groups. With `plyranges`
it is easy to filter DNase sites by overlap (or lack of overlap)
with JUN peaks:

```{r message=FALSE, warning=FALSE}
## Define focal and pool
focal <- dnase |>
  filter_by_overlaps(junPeaks)

pool <- dnase |>
  filter_by_non_overlaps(junPeaks)

```

The focal set must be smaller than the pool set for matching
to work correctly. Matching is most effective when the pool
set is much larger and covers all values in the focal set.

```{r message=FALSE, warning=FALSE}
length(focal)
length(pool)

length(pool)/length(focal)
```

Before matching, the focal set shows a different
distribution of length and signal than the pool 
set:

```{r, message=FALSE, warning=FALSE}
## Before matching, focal shows higher
## signalValue than pool
library(tidyr)
library(ggplot2)
bind_ranges(focal=focal,
            pool=pool,
            .id="set") |>
  as.data.frame() |>
  pivot_longer(cols=c("log10_length", "log10_signal"),
               names_to="features",
               values_to="values") |>
  ggplot(aes(values, color=set)) +
  facet_wrap(~features) +
  stat_density(geom='line', position='identity') +
  ggtitle("DNase sites") +
  theme_bw() +
  theme(plot.title=element_text(hjust=0.5))
```

## Obtaining the matched set with `matchRanges()`

To control for these differences, we can use `matchRanges()` to
select a subset of unbound DNase sites from the pool that have
the same distribution of length and signal.

```{r, message=FALSE, warning=FALSE}
library(nullranges)
set.seed(123)
mgr <- matchRanges(focal=focal,
                   pool=pool,
                   covar=~log10_length + log10_signal,
                   method='rejection',
                   replace=FALSE)

mgr
```

## Assessing covariate balance

Now let's use the `plotCovariate()` function with `patchwork` to
visualize how similar our matched distribution is to focal:

```{r, message=FALSE, warning=FALSE}
library(patchwork)
lapply(covariates(mgr),
       plotCovariate,
       x=mgr,
       sets=c('f', 'm', 'p')) |>
  Reduce('+', x=_) +
  plot_layout(guides="collect") +
  plot_annotation(title="DNase sites",
                  theme=theme(plot.title=element_text(hjust=0.40)))
```

An important part of propensity-score matching, is assessing
similarity, or balance, between the focal and matched sets.
One way is to visually examine the distributions as we have
done above. Another way is to report summary statistics about
the two sets. `cobalt` is a package designed to specifically
address covariate balance after covariate matching. Below, we
demonstrate how to use `cobalt` to calculate the standardized mean
differences and visualize these statistics with a love plot.
For more information about assessing covariate balance, refer
to the detailed documentation in the `cobalt` vignette:
`vignette("cobalt", package = "cobalt")`.

```{r, message=FALSE, warning=FALSE}
library(cobalt)
res <- bal.tab(f.build("set", covariates(mgr)),
               data=matchedData(mgr)[!set %in% 'unmatched'],
               distance="ps",
               focal="focal",
               which.treat="focal",
               s.d.denom="all")

res

plot(res) + xlim(c(-2, 2))

```

The "focal vs. matched" comparison shows much lower
standardized mean differences than "focal vs. pool",
indicating that the matched set has been
successfully controlled for covariates of DNAse site
length and signal.