---
title: "Matching with GRanges"
author: "Eric S. Davis"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
output:
  rmarkdown::html_document:
    highlight: tango
    toc: true
    toc_float: true	
    fig_width: 5
    fig_height: 3
vignette: |
  %\VignetteIndexEntry{Using matchRanges}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

In this vignette we demonstrate generating covariate-matched, null-hypothesis GRanges using the `matchRanges()` function to test for the enrichment of CCCTC-binding factor (CTCF) at chromatin loop anchors. Additionally, we provide usage examples for accessor functions and covariate visualization for matched datasets.

## Background and motivation

One of the fundamental principles of chromatin-looping suggests that most loops are bound at both ends by the CTCF transcription factor (TF). CTCF-bound loops can be formed by loop-extrusion, where the ring-like cohesin complex extrudes chromatin until stopped by a bound CTCF TF. By this mechanism, we expect most loop anchors will be bound by CTCF.

While we could test this hypothesis by simple overlap or permutation testing, these approaches fail to account for non-randomly distributed covariate genomic features. For example, loop anchors are commonly bound by CTCF and located in open chromatin regions. We can use `matchRanges()` to test for CTCF enrichment at loop anchors independently from open chromatin regions.

Here, we generate a set of null-hypothesis GRanges to more rigorously test CTCF enrichment at loop anchors independently from open chromatin regions. We use the `hg19_10kb_bins` dataset from the `nullrangesData` package, which contains ranges for every 10Kb bin along the genome with CTCF, DNase, and loop feature annotations from GM12878 (see `?nullrangesData::hg19_10kb_bins`).

## Matching with `matchRanges()`

Before we generate our null ranges, let's take a look at our example dataset:

```{r, message=FALSE, warning=FALSE}
library(nullrangesData)

## Load example data
data("hg19_10kb_bins")
bins <- hg19_10kb_bins

bins
```

`matchRanges()` works by selecting a set of covariate-matched controls from a pool of options based on an input focal set of interest. Here, we define `focal` as bins that contain a loop anchor, `pool` as bins that don't contain a loop anchor, and `covar` as DNase signal and number of DNase sites per bin:

```{r}
## Define focal, pool, and covar
focal <- bins[bins$looped]
pool  <- bins[!bins$looped]
covar <- ~dnaseSignal + n_dnase_sites
```

Then we can implement matching using either nearest-neighbor matching (`method = 'nearest'`), rejection sampling (`method = 'rejection'`), or stratified sampling (`method = 'stratified'`), with or without replacement (`replace = TRUE/FALSE`).

```{r, message=FALSE}
library(nullranges)

system.time({
  ## Match ranges
  set.seed(123)
  mgr <- matchRanges(focal = focal,
                     pool = pool,
                     covar = covar,
                     method = 'rejection',
                     replace = FALSE)  
})

mgr
```

When the focal and pool arguments are `GRanges` objects, `matchRanges()` returns a `MatchedGRanges` object. The `MatchedGRanges` class extends `GRanges`, so all of the same operations can be applied:

```{r, message=FALSE}
library(GenomicRanges)
library(plyranges)

## Resize operation on MatchedGRanges object
mgr %>% resize(width = 100, fix = 'center')

## Summarize ctcfSignal by n_ctcf_sites
mgr %>%
  group_by(n_ctcf_sites) %>%
  summarize(mean(ctcfSignal))
```

## Matching quality

We can get a quick summary of the matching quality with `overview()`:

```{r}
overview(mgr)
```

For continuous covariates (such as `dnaseSignal`), `overview()` shows the mean and standard deviation between each matched set. For factor covariates, such as `n_dnase_sites`, `overview()` reports the number of observations in factor level and matched set. The bottom section shows the mean and s.d (or n, for factors) difference between focal and matched sets.

`overview()` also summarizes the propensity scores for each set to give a quick idea of overall matching quality.

### Visualizing propensity scores

To visualize overall propensity-score matching quality, use `plotPropensity():`

```{r}
plotPropensity(mgr)
```

The default plotting method is `'ridges'` but the `type` argument can be used to plot `'lines'` or `'jitter'`. Using `type = 'jitter'` is discouraged on large datasets because the large density of points can stall the R-graphics device.

```{r}
plotPropensity(mgr, type = 'lines')
```

### Visualizing covariate distributions

To determine matching quality, we can visualize covariate distributions across each of the matched sets. The `covariates()` function accesses the matched covariates from the `MatchedGRanges` object:

```{r}
covariates(mgr)
```

Distributions can be visualized with `plotCovariates()`:

```{r}
plotCovariates(mgr, type = 'lines', covar = covariates(mgr)[1])
```

TODO: Add more information after `plotCovariate()` is finished

### Extracting matched data

The matched data is easily accessible for users that wish to make custom plots of either propensity score or covariate distributions:

```{r}
matchedData(mgr)
```

## Accessing matched sets

`nullranges` provides convenience accessor functions for extracting matched sets from `MatchedGRanges` objects:

```{r}
focal(mgr)
pool(mgr)
matched(mgr)
unmatched(mgr)
```

As well as extracting indices by group:

```{r}
indices(mgr, group = 'matched') %>% head()
```

The default group is `'matched'`, but `indices` can be used to pull from all sets. Note that the indices for `group = 'focal'` come from the focal set, whereas indices for `group = 'pool'` , `group = 'matched'`, and `group = 'unmatched'` come from the pool set:

```{r}
## All should be TRUE
identical(focal(mgr), focal(mgr)[indices(mgr, group = 'focal')])
identical(pool(mgr), pool(mgr)[indices(mgr, group = 'pool')])
identical(matched(mgr), pool(mgr)[indices(mgr, group = 'matched')])
identical(unmatched(mgr), pool(mgr)[indices(mgr, group = 'unmatched')])
```

## Compare CTCF sites

Using our matched ranges, we can compare enrichment of CTCF between bins that 1) contain a loop anchor (i.e. looped), 2) don't contain a loop anchor (i.e. unlooped), or 3) don't contain a loop anchor, but are also matched for the strength and number of DNase sites (i.e. matched). In this case, we calculate CTCF enrichment as the percent of bins that contain CTCF among our 3 groups:

```{r fig.width=4, fig.height=4}
## Percent of bins with CTCF
g1 <- (sum(focal(mgr)$n_ctcf_sites >= 1) / length(focal(mgr))) * 100
g2 <- (sum(pool(mgr)$n_ctcf_sites >= 1) / length(pool(mgr))) * 100
g3 <- (sum(mgr$n_ctcf_sites >= 1) / length(mgr)) * 100

## Visualize
barplot(height = c(g1, g2, g3),
        names = c('looped', 'unlooped', 'matched'),
        ylab = "CTCF occupied bins (%)",
        col = c('#1F78B4', '#33A02C', '#A6CEE3'),
        main = 'CTCF enrichment')
```
