---
title: "Segmented block bootstrap"
author: "Wancen Mu"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
output:
  rmarkdown::html_document:
    highlight: tango
    toc: true
    toc_float: true	
vignette: |
  %\VignetteIndexEntry{Segmented block bootstrap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction to segmented block bootstrap

cite Bickel et al...

Options:

* Segment genome using feature density, e.g. gene density
* Use exiting segmentation (e.g. ChromHMM, etc.)

# Segmentation by gene density

First we obtain the Ensembl genes ...

```{r}
library(EnsDb.Hsapiens.v86)
edb <- EnsDb.Hsapiens.v86
g <- genes(edb, filter = AnnotationFilterList(GeneIdFilter("ENSG", "startsWith")))
```

```{r}
library(GenomeInfoDb)
g <- keepStandardChromosomes(g, pruning.mode = "coarse", species = "Homo_sapiens")
seqlevelsStyle(g) <- "UCSC"
g <- sort(g)
g <- keepSeqlevels(g, as.character(seq_len(9)), pruning.mode = "coarse")
```

## Import deny region

Deny regions ... cite Kundaje

```{r}
library(nullrangesData)
data("deny")
```

```{r}
library(nullranges)
library(plyranges)
```

## CBS segmentation

We subset the deny ranges to those which are 500 bp or larger.

cite: DNAcopy paper

```{r}
deny2 <- deny %>% filter(width(deny) >= 500)
L_s <- 1e6
seg <- segment_density(gr, n = 3, L_s = L_s,
                       deny = deny2, type = "cbs",
                       boxplot = TRUE)
```

## Alternatively: HMM segmentation

cite: RcppHMM

```{r}
seg_hmm <- segment_density(gr, n = 3, L_s = L_s,
                           deny = deny2, type = "hmm",
                           boxplot = TRUE)
```

# Segmented block bootstrap within chromosome

Here we apply a segmented block bootstrap with blocks of size 500kb,
to the genes...

Test on 100 times bootstrap ~17s

```{r}
L_b <- 5e5
system.time({
  res_wochr <- seg_bootstrap_granges(seg, gr, L_b, R = 100,
                                     within_chrom = FALSE, proportion_length = FALSE,
                                     coarse = TRUE)
})
system.time({
  res_wochr_propl <- seg_bootstrap_granges(seg, gr, L_b, R = 100,
                                           within_chrom = FALSE, coarse = TRUE)
})
system.time({
  res_wchr <- seg_bootstrap_granges(seg, gr, L_b, R = 100,
                                    within_chrom = TRUE, proportion_length = FALSE,
                                    coarse = TRUE)
})
system.time({
  res_wchr <- seg_bootstrap_granges(seg, gr, L_b, R = 100,
                                    within_chrom = TRUE, proportion_length = TRUE,
                                    coarse = TRUE)
  })
```

# Test performance on grid of segmentation and block lengths

TODO: this may not be ideal for the vignette, but better for a paper

```{r eval=FALSE}
Ls <- c(1e5, 1e6, 5e6)
L_b <- c(5e4, 5e5, 1e6)
mapply(function(Ls, L_b) {
  seg <- segment_density(g1, n = 3, Ls = Ls, type = "CBS", plot_origin = TRUE)
  seg_length <- seg %>%
    group_by(state) %>%
    summarise(Ls = sum(width)) # derive each states length
  L_c <- seqlengths(g1)["9"]
  L_s <- seg_length$Ls
  system.time(res <- seg_bootstrap_iranges(seg, g1, L_c, L_s, L_b))[[3]]
  res <- do.call(c, res)
  res <- sort(res)
  CBS_boot <- GRanges(seqnames = "9", ranges = res, seqlengths = L_c)
  counts_cbs_boot <- countOverlaps(query, CBS_boot)
  eps <- rnorm(length(counts_cbs_boot), 0, .2)
  print(hist(counts_cbs_boot, breaks = 50, xlab = paste(Ls, "with", L_b, "block size")))
  print(plot(sqrt(counts_cbs_boot) + eps))
}, Ls, L_b)
```

# Segmented block bootstrap across chromosome

Toy example for evaluating segmented block bootstrap...

```{r}
library(GenomicRanges)
seg <- GRanges("1", IRanges(c(1, 101, 201, 301),
  c(100, 200, 300, 400),
  state = c(1:3, 1)
))
gr <- GRanges("1",
  IRanges(c(21, 26, 31, 36, 41, 221, 231, 241), width = 4),
  seqlengths = c("1" = 400)
  )
gr$state <- seg$state[ findOverlaps(x, seg, select="first") ]
```

```{r}
suppressPackageStartupMessages(library(BentoBox))
plotGRanges <- function(gr) {
  p <- bb_params(chromstart = 0, chromend = 500,
                 x = 0.5, width = 4, height = 0.5,
                 at = seq(0, 500, 50),
                 fill = palette()[2:4])
  bb_pageCreate(width = 5, height = 2, xgrid = 0,
                ygrid = 0, showGuides = FALSE)
  for (i in seq_along(seqlevels(gr))) {
    chrom <- seqlevels(gr)[i]
    chromend <- seqlengths(gr)[[chrom]]
    suppressMessages({
      pbed <- bb_plotBed(data = gr, params = p,
                         chrom = chrom,
                         y = 0.25 + (i-1)*.7,
                         just = c("left", "bottom"),
                         colorby=colorby("state"))
      bb_annoGenomeLabel(plot = pbed, params = p, y = 0.30 + (i-1)*.7)
    })
  }
}
```

```{r}
plotGRanges(x)
```
